---
title: "Modeling Physiological and Behavioral Predictors of Stress Using a Linear Mixed Model on Wearable Health Data"
subtitle: "Using Synthetic Longitudinal Wearable Device Data"
author: "Savannah Davis<br>Bryce Mattos<br>Jessica Jones<br>Jordan Gainey<br>(Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"}


## Introduction

The introduction should:

-   Develop a storyline that captures attention and maintains interest.

-   Your audience is your peers

-   Clearly state the problem or question you're addressing.

-   Introduce why it is relevant needs.

-   Provide an overview of your approach.

    Stress is a state where the emotional, psychological, or physiological systems of an individual experience a demand that exceeds their knowledge and ability to resolve. Whether perceived consciously or unconsciously, the body then responds by prompting organs and glands to produce hormones that will assist the individual in adapting to the situation. Normal physiological responses to acute stress are short-lived and allow the body to return to its natural state without any long-term consequences. However, chronic stress, which is the result of unresolved long-standing demands and challenges can result in physical complications and disease in the body [@can2022survey]. In fact, the effects of chronic stress are so serious that research in Europe showed stress as the second most severe work-related health issue [@can2022survey]. Insights on stress can of course be self-reported by individuals, however, more efficient and accurate means of tracking stress can be accomplished via wearable devices such as Fitbit, Garmin, etc.

	  Wearable devices can collect a wide range of data such as: heart rate, heart rate variability(HRV), sleep quality, physical activity, and movement patterns [@namvari2022ppg; @wang2024heart]. Out of those variables, researchers select the physiological ones that are directly influenced by the autonomic nervous system (ANS), mainly because stress is fundamentally an ANS response. The ANS regulates involuntary bodily functions and also consists of the sympathetic(SNS) and parasympathetic nervous systems(PNS) which maintain homeostasis within the body [@chalmers2022sleep]. During periods of stress, the SNS activates which results in elevated heart rates and changes in HRV, which makes those the perfect variables to include in stress prediction research. Additionally, sleep, physical activity, circadian rhythms, and environmental conditions also contribute to stress and researchers often incorporate them when developing stress prediction models[@wang2024heart].
	
	  While physiological variables such as heart rate and HRV provide valuable information into how the body responds to stress, researchers have found that behavioral and lifestyle factors also contribute to stress levels. Sleep duration, sleep quality, sleep efficiency, screen time, caffeine consumption, alcohol use, and mindfulness practices all have been shown to affect overall well-being [@vidalbustamante2024sleep; @booth2022stress]. Out of those listed factors, sleep has been identified as one of the strongest predictors of perceived stress, as disruptions influence an individuals ability to adapt to daily demands [@bloomfield2024sleep]. So, to increase the accuracy of stress prediction, researchers have concluded that combining physiological signals with lifestyle factors improves the ability to predict persons experiencing higher levels of stress [@sano2018markers].
  
    Our dataset consists of the following:
-   300 users observed over a six-month period.

-   55,200 total observations

-   Variables: HRV, sleep duration, sleep efficiency, caffeine consumption, alcohol use, screen time, mindfulness, and physical activity, and their association with stress scores over time. 

  
    The statistical method chosen to aid in the prediction of stress using our simulated wearable device data is the linear mixed model. This method suits our dataset well because it contains repeated measurements from the same participants over time, and linear mixed models are specifically tailored to longitudinal data, trend identification, repeated observations, and variation between individuals [@silveira2023mixedeffects; @harrison2018mixed].



## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.


This report utilizes the *Health + Wellness + Stress/Sleep Tracking* dataset obtained from Kaggle. The dataset contains wearable health measurements and behavioral measurements for 300 simulated individuals over a 6 month period. Variables include but are not limited to: demographic characteristics, physical activity, sleep metrics, cardiovascular measurements, and stress score. This dataset is longitudinal and is suitable for analyzing physiological and behavioral predictors of stress using a linear mixed model. [@kaggle_dataset]

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(corrplot)
library(lme4)
library(lmerTest)
library(performance)
library(broom.mixed)
library(car)
```

```{r}
# Load Data

df = read_csv("wearables_health_6mo_daily.csv")
```

```{r}
# Quick overview of dataset

dataset_overview <- tibble(
  Metric = c("Total Observations", "Total Variables", "Unique Users"),
  Value = c(
    scales::comma(nrow(df)),
    scales::comma(ncol(df)),
    scales::comma(n_distinct(df$user_id))
  )
)

dataset_overview %>%
  kable(caption = "Dataset Overview") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

The raw dataset contains over 55k records for 300 distinct users, with over 30 variables to choose to build the model from.

#### Descriptive Statistics

```{r}
# Function to create descriptive statistics tables

create_descriptive_table <- function(data, variables, caption) {

  data %>%
    select(all_of(variables)) %>%
    pivot_longer(
      cols = everything(),
      names_to = "Variable",
      values_to = "Value"
    ) %>%
    group_by(Variable) %>%
    summarise(
      Mean = round(mean(Value, na.rm = TRUE), 2),
      SD = round(sd(Value, na.rm = TRUE), 2),
      Min = round(min(Value, na.rm = TRUE), 2),
      Median = round(median(Value, na.rm = TRUE), 2),
      Max = round(max(Value, na.rm = TRUE), 2),
      .groups = "drop"
    ) %>%
    kable(caption = caption) %>%
    kable_styling(
      bootstrap_options = c(
        "striped",
        "hover",
        "condensed"
      ),
      full_width = FALSE
    )
}
```

The variables in the dataset are split into several different categories to assist in choosing variables of interest. Descriptive statistics tables will be created for each category including the mean, standard deviation, 
minimum, median, and maximum values.

```{r}
# Demographic fields

create_descriptive_table(
  df,
  c(
    "age",
    "height_cm",
    "weight_kg",
    "bmi"
  ),
  "Demographic Characteristics"
)
```


```{r}
# Physiological fields

create_descriptive_table(
  df,
  c(
    "resting_hr_bpm",
    "avg_hr_day_bpm",
    "hrv_rmssd_ms",
    "spo2_avg_pct",
    "sbp_mmHg",
    "dbp_mmHg"
  ),
  "Physiological Characteristics"
)
```

```{r}
# Behavioral fields

create_descriptive_table(
  df,
  c(
    "steps",
    "distance_km",
    "calories_kcal",
    "workout_minutes",
    "caffeine_mg",
    "alcohol_units",
    "screen_time_min",
    "mindfulness_minutes"
  ),
  "Behavioral Characteristics"
)
```

```{r}
# Sleep fields

create_descriptive_table(
  df,
  c(
    "sleep_duration_hours",
    "sleep_efficiency",
    "sleep_latency_min",
    "wake_after_sleep_onset_min",
    "sleep_stage_rem_pct",
    "sleep_stage_deep_pct",
    "sleep_stage_light_pct"
  ),
  "Sleep Characteristics"
)
```

```{r}
# Outcome variable

create_descriptive_table(
  df,
  c(
    "stress_score"
  ),
  "Stress Score Summary"
)
```

#### Missing Values

The dataset is investigated for missing values to assess next steps for preprocessing.

```{r}
# Investigating missing values

missing_values <- df %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Missing_Values"
  ) %>%
  mutate(
    Percent_Missing = round(100 * Missing_Values / nrow(df), 2)
  ) %>%
  filter(
    Percent_Missing > 0
  ) %>%
  arrange(desc(Missing_Values)) %>%
  mutate(
    Missing_Values = scales::comma(Missing_Values)
  )

missing_values %>%
  kable(caption = "Missing Values by Variable") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

```{r}
# Count rows containing at least one missing value

rows_with_missing <- sum(!complete.cases(df))

# Calculate percentage of dataset removed

pct_rows_removed <- round(
  100 * rows_with_missing / nrow(df),
  2
)

missing_summary <- tibble(
  Total_Observations = scales::comma(nrow(df)),
  Rows_With_Missing = scales::comma(rows_with_missing),
  Percent_Removed = paste0(pct_rows_removed, "%")
)

missing_summary %>%
  kable(
    caption = "Impact of Missing Data Removal"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

It is revealed that less than 2% of the dataset contains missing values across categories such as the physiological, behavioral, and sleep characteristic fields.

#### Categorical Fields

```{r}
# Function to graph frequencies of categorical data

plot_categorical <- function(data, column) {

  plot_data <- data %>%
    count({{ column }}) %>%
    mutate(
      percent = round(100 * n / sum(n), 1),
      label = paste0(
        scales::comma(n),
        "\n(",
        percent,
        "%)"
      )
    )

  ggplot(
    plot_data,
    aes(
      x = reorder(as.character({{ column }}), -n),
      y = n
    )
  ) +
    geom_col() +
    geom_text(
      aes(label = label),
      vjust = -0.3,
      size = 3.5
    ) +
    scale_y_continuous(
      expand = expansion(mult = c(0, 0.15))
    ) +
    labs(
      title = paste(
        "Frequency of",
        tools::toTitleCase(
          gsub("_", " ", deparse(substitute(column)))
        )
      ),
      x = NULL,
      y = "Count"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(
        face = "bold",
        hjust = 0.5
      ),
      axis.text.x = element_text(
        angle = 30,
        hjust = 1
      )
    )
}
```

The categorical fields are explored to determine the frequencies in each category.

```{r}
# Apply graphing function

plot_categorical(df, mood)

plot_categorical(df, gender)

plot_categorical(df, region)

plot_categorical(df, workout_type)

plot_categorical(df, device_model)
```

#### Stress Distribution

The outcome field is explored to determine distribution.

```{r}
# Investigating distribution of stress score within the dataset

ggplot(df, aes(x = stress_score)) +
  geom_histogram(bins = 20) +
  labs(
    title = "Distribution of Stress Scores",
    subtitle = "Stress scores appear approximately normally distributed",
    x = "Stress Score",
    y = "Count"
  ) +
  theme_minimal()
```


### Modeling


```{r}
# Remove null values since there aren't many
df_clean <- na.omit(df)

# Convert categorical data to factor
df_clean <- df_clean %>%
  mutate(
    user_id = as.factor(user_id),
    gender = as.factor(gender),
    region = as.factor(region),
    device_model = as.factor(device_model),
    workout_type = as.factor(workout_type),
    mood = as.factor(mood)
  )

# Select all categorical variables
categorical_vars <- df_clean %>%
  select(
    user_id,
    gender,
    region,
    device_model,
    workout_type,
    mood
  )

# Select all continuous variables
continuous_vars <- df_clean %>%
  select(
    age, height_cm, weight_kg, bmi, resting_hr_bpm,
    avg_hr_day_bpm, hrv_rmssd_ms, spo2_avg_pct, sbp_mmHg,
    dbp_mmHg, steps, distance_km, calories_kcal,
    workout_minutes, caffeine_mg, alcohol_units, screen_time_min, 
    sleep_duration_hours, sleep_efficiency, sleep_latency_min,
    wake_after_sleep_onset_min, sleep_stage_rem_pct, sleep_stage_deep_pct, 
    sleep_stage_light_pct, stress_score, mindfulness_minutes
  )
```


```{r}
# Calculate correlation matrix for continuous variables
# Helps identify correlated variables before building model
corr_matrix <- cor(
  continuous_vars,
  use = "complete.obs"
)

# Correlation heatmap
corrplot(
  corr_matrix,
  method = "color",
  type = "upper",
  tl.cex = 0.7
)
```

The correlation heatmap reveals several fields that need to be investigated further to determine the strength of the correlation. If certain fields are highly correlated, then one may be chosen to be left out of the model to reduce redundancy.

```{r}
# Create table from the corr matrix
corr_table <- as.data.frame(as.table(corr_matrix)) %>%
  rename(
    Variable1 = Var1,
    Variable2 = Var2,
    Correlation = Freq
  ) %>%
  filter(
    as.character(Variable1) < as.character(Variable2)
  ) %>%
  mutate(
    abs_corr = abs(Correlation)
  ) %>%
  arrange(desc(abs_corr))

# Filter to only the highly correlated pairs and remove duplicate pairs
corr_table %>%
  filter(abs_corr >= 0.70) %>%
  mutate(
    Correlation = round(Correlation, 2)
  ) %>%
  select(
    Variable1,
    Variable2,
    Correlation
  ) %>%
  kable(
    caption = "Strong correlations among predictor variables"
  ) %>%
  kable_styling(full_width = FALSE)
```

The steps and distance fields are highly correlated which makes sense. Another obvious pair is the bmi and weight fields. Interestingly, the steps and deep sleep stage percentage fields are strongly correlated.


The first model is built which includes all fields except for the pairs of strongly correlated fields.

```{r}
# Build initial model with all variables (except ones with strong correlations)
# Put user_id as the random effect

lmm_full <- lmer(
  stress_score ~
    age +
    gender +
    region +
    device_model +
    workout_type +
    bmi +
    resting_hr_bpm +
    hrv_rmssd_ms +
    spo2_avg_pct +
    sbp_mmHg +
    dbp_mmHg +
    distance_km +
    workout_minutes +
    caffeine_mg +
    alcohol_units +
    screen_time_min +
    mindfulness_minutes +
    sleep_duration_hours +
    sleep_efficiency +
    sleep_latency_min +
    wake_after_sleep_onset_min +
    (1 | user_id),
  data = df_clean
)

summary(lmm_full)
```


Multicollinearity is tested using VIF.

```{r}
# Test mutlicollinearity using VIF

vif_full <- vif(
  lm(
    stress_score ~
      age +
      gender +
      region +
      device_model +
      workout_type +
      bmi +
      resting_hr_bpm +
      hrv_rmssd_ms +
      spo2_avg_pct +
      sbp_mmHg +
      dbp_mmHg +
      distance_km +
      workout_minutes +
      caffeine_mg +
      alcohol_units +
      screen_time_min +
      mindfulness_minutes +
      sleep_duration_hours +
      sleep_efficiency +
      sleep_latency_min +
      wake_after_sleep_onset_min,
    data = df_clean
  )
)

vif_full
```



```{r}
# Test model after taking out region and device_model because they were not meaningful

lmm_no_region_device <- lmer(
  stress_score ~
    age +
    gender +
    workout_type +
    bmi +
    resting_hr_bpm +
    hrv_rmssd_ms +
    spo2_avg_pct +
    sbp_mmHg +
    dbp_mmHg +
    distance_km +
    workout_minutes +
    caffeine_mg +
    alcohol_units +
    screen_time_min +
    mindfulness_minutes +
    sleep_duration_hours +
    sleep_efficiency +
    sleep_latency_min +
    wake_after_sleep_onset_min +
    (1 | user_id),
  data = df_clean
)

summary(lmm_no_region_device)
```


```{r}
# Reduced model version 1 after getting rid of irrelevant variables
# Variables with weak effects and little explanatory value were removed

lmm_reduced <- lmer(
  stress_score ~
    workout_type +
    hrv_rmssd_ms +
    distance_km +
    caffeine_mg +
    alcohol_units +
    screen_time_min +
    mindfulness_minutes +
    sleep_duration_hours +
    sleep_efficiency +
    wake_after_sleep_onset_min +
    (1 | user_id),
  data = df_clean
)

summary(lmm_reduced)
```

```{r}
# Test taking out workout_type to see if it contributes to model fit
lmm_reduced_2 <- lmer(
  stress_score ~
    hrv_rmssd_ms +
    distance_km +
    caffeine_mg +
    alcohol_units +
    screen_time_min +
    mindfulness_minutes +
    sleep_duration_hours +
    sleep_efficiency +
    wake_after_sleep_onset_min +
    (1 | user_id),
  data = df_clean
)

# Refit both models with maximum likelihood estimation (ML)
lmm_reduced_ml <- update(
  lmm_reduced,
  REML = FALSE
)

lmm_reduced_2_ml <- update(
  lmm_reduced_2,
  REML = FALSE
)

# Run anova
anova(
  lmm_reduced_2_ml,
  lmm_reduced_ml
)
```

```{r}
# Test removing wake_after_sleep_onset_min to see if it contributes to model fit

lmm_reduced_3 <- lmer(
  stress_score ~
    workout_type +
    hrv_rmssd_ms +
    distance_km +
    caffeine_mg +
    alcohol_units +
    screen_time_min +
    mindfulness_minutes +
    sleep_duration_hours +
    sleep_efficiency +
    (1 | user_id),
  data = df_clean
)

# Refit model with maximum likelihood estimation (ML)
lmm_reduced_3_ml <- update(
  lmm_reduced_3,
  REML = FALSE
)

# Run anova
anova(
  lmm_reduced_3_ml,
  lmm_reduced_ml
)
```


```{r}
# Build final model
# The final predictor set was determined through iterative model reduction

lmm_final <- lmer(
  stress_score ~
    workout_type +
    hrv_rmssd_ms +
    distance_km +
    caffeine_mg +
    alcohol_units +
    screen_time_min +
    mindfulness_minutes +
    sleep_duration_hours +
    sleep_efficiency +
    wake_after_sleep_onset_min +
    (1 | user_id),
  data = df_clean
)

summary(lmm_final)
```


```{r}
# Evaluate model assumptions and overall fit
plot(lmm_final)

qqnorm(residuals(lmm_final))
qqline(residuals(lmm_final))
```

```{r}
# The random intercepts should be approximately normally distributed
qqnorm(ranef(lmm_final)$user_id[,1])
qqline(ranef(lmm_final)$user_id[,1])
```

```{r}
# Check model diagnostics
check_model(lmm_final, check = "qq")
check_model(lmm_final, check = "homogeneity")
check_model(lmm_final, check = "linearity")
check_model(lmm_final, check = "outliers")
```


```{r}
# Create an 80/20 train-test split at the user level
# Splitting by user prevents observations from the same participant appearing in both datasets

set.seed(1021)

users <- unique(df_clean$user_id)

train_users <- sample(
  users,
  size = round(length(users) * 0.80)
)

# Assign observations from training users to the training dataset
train_data <- df_clean %>%
  filter(user_id %in% train_users)

# Assign observations from held-out users to the testing dataset
test_data <- df_clean %>%
  filter(!user_id %in% train_users)

# Verify the number of unique users in each partition
length(unique(train_data$user_id))
length(unique(test_data$user_id))

# Verify the number of observations in each partition
nrow(train_data)
nrow(test_data)
```

```{r}
# Check to make sure there are only unique users in train and test split
length(
  intersect(
    unique(train_data$user_id),
    unique(test_data$user_id)
  )
)
```

```{r}
# Use the final model from before and build it from train dataset
lmm_train <- lmer(
  stress_score ~
    workout_type +
    hrv_rmssd_ms +
    distance_km +
    caffeine_mg +
    alcohol_units +
    screen_time_min +
    mindfulness_minutes +
    sleep_duration_hours +
    sleep_efficiency +
    wake_after_sleep_onset_min +
    (1 | user_id),
  data = train_data
)
```


```{r}
# Predict on our test data
test_pred <- predict(
  lmm_train,
  newdata = test_data,
  allow.new.levels = TRUE
)
```

```{r}
# Evaluate predictive performance on the test dataset

actual <- test_data$stress_score

rmse <- sqrt(mean((actual - test_pred)^2))

mae <- mean(abs(actual - test_pred))

r2_test <- cor(actual, test_pred)^2
```

### Results

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

**Add results discussion here**

```{r}
# Plot predicted vs actual stress score to measure performance of model
ggplot(
  data.frame(actual, test_pred),
  aes(x = actual, y = test_pred)
) +
  geom_point(
    alpha = 0.05,
    size = 1
  ) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed"
  ) +
  labs(
    title = "Predicted vs Actual Stress Scores",
    x = "Actual Stress Score",
    y = "Predicted Stress Score"
  ) +
  theme_minimal()
```


```{r}
# This table summarizes the statistical significance of each predictor retained in the final model

coef_table <- broom.mixed::tidy(lmm_final)

coef_table %>%
  filter(effect == "fixed") %>%
  select(
    term,
    estimate,
    std.error,
    statistic,
    p.value
  ) %>%
  mutate(
    across(
      c(estimate, std.error, statistic),
      ~ round(.x, 3)
    ),
    p.value = ifelse(
      p.value < .001,
      "<0.001",
      round(p.value, 3)
    )
  ) %>%
  kable(
    caption = "Fixed Effects from Final Linear Mixed Model"
  ) %>%
  kable_styling(
    bootstrap_options = c(
      "striped",
      "hover",
      "condensed"
    )
  )
```

```{r}
# Calculate final model performance metrics

r2_vals <- r2(lmm_final)
icc_val <- icc(lmm_final)

performance_table <- tibble(
  Metric = c(
    "Marginal R²",
    "Conditional R²",
    "ICC",
    "AIC",
    "Users",
    "Observations"
  ),
  Value = c(
    round(r2_vals$R2_marginal, 3),
    round(r2_vals$R2_conditional, 3),
    round(icc_val$ICC_adjusted, 3),
    round(AIC(lmm_final), 1),
    length(unique(df_clean$user_id)),
    nrow(df_clean)
  )
)

performance_table %>%
  kable(
    caption = "Final Model Performance Metrics"
  ) %>%
  kable_styling(
    bootstrap_options = c(
      "striped",
      "hover",
      "condensed"
    )
  )
```

```{r}
# Summarize predictive performance on the test dataset

prediction_table <- tibble(
  Metric = c(
    "Test R²",
    "RMSE",
    "MAE"
  ),
  Value = c(
    round(r2_test, 3),
    round(rmse, 3),
    round(mae, 3)
  )
)

prediction_table %>%
  kable(
    caption = "Prediction Performance on Test Data"
  ) %>%
  kable_styling(
    bootstrap_options = c(
      "striped",
      "hover",
      "condensed"
    )
  )
```


### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

**Add conclusion here**

## References
