---
title: "Modeling Physiological and Behavioral Predictors of Stress Using a Linear Mixed Model on Wearable Health Data"
subtitle: "Using Synthetic Longitudinal Wearable Device Data"
author: "Savannah Davis<br>Bryce Mattos<br>Jessica Jones<br>Jordan Gainey<br>(Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"}


## Introduction

The introduction should:

-   Develop a storyline that captures attention and maintains interest.

-   Your audience is your peers

-   Clearly state the problem or question you're addressing.

-   Introduce why it is relevant needs.

-   Provide an overview of your approach.

**Add intro here**


## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.


This report utilizes the *Health + Wellness + Stress/Sleep Tracking* dataset obtained from Kaggle. The dataset contains wearable health measurements and behavioral measurements for 300 simulated individuals over a 6 month period. Variables include but are not limited to: demographic characteristics, physical activity, sleep metrics, cardiovascular measurements, and stress score. This dataset is longitudinal and is suitable for analyzing physiological and behavioral predictors of stress using a linear mixed model. [@kaggle_dataset]

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(corrplot)
library(lme4)
library(lmerTest)
```

```{r}
# Load Data

df = read_csv("wearables_health_6mo_daily.csv")
```

```{r}
# Quick overview of dataset

dataset_overview <- tibble(
  Metric = c("Total Observations", "Total Variables", "Unique Users"),
  Value = c(
    scales::comma(nrow(df)),
    scales::comma(ncol(df)),
    scales::comma(n_distinct(df$user_id))
  )
)

dataset_overview %>%
  kable(caption = "Dataset Overview") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

#### Descriptive Statistics

```{r}
# Function to create descriptive statistics tables

create_descriptive_table <- function(data, variables, caption) {

  data %>%
    select(all_of(variables)) %>%
    pivot_longer(
      cols = everything(),
      names_to = "Variable",
      values_to = "Value"
    ) %>%
    group_by(Variable) %>%
    summarise(
      Mean = round(mean(Value, na.rm = TRUE), 2),
      SD = round(sd(Value, na.rm = TRUE), 2),
      Min = round(min(Value, na.rm = TRUE), 2),
      Median = round(median(Value, na.rm = TRUE), 2),
      Max = round(max(Value, na.rm = TRUE), 2),
      .groups = "drop"
    ) %>%
    kable(caption = caption) %>%
    kable_styling(
      bootstrap_options = c(
        "striped",
        "hover",
        "condensed"
      ),
      full_width = FALSE
    )
}
```

```{r}
# Demographic fields

create_descriptive_table(
  df,
  c(
    "age",
    "height_cm",
    "weight_kg",
    "bmi"
  ),
  "Demographic Characteristics"
)
```

```{r}
# Physiological fields

create_descriptive_table(
  df,
  c(
    "resting_hr_bpm",
    "avg_hr_day_bpm",
    "hrv_rmssd_ms",
    "spo2_avg_pct",
    "sbp_mmHg",
    "dbp_mmHg"
  ),
  "Physiological Characteristics"
)
```

```{r}
# Behavioral fields

create_descriptive_table(
  df,
  c(
    "steps",
    "distance_km",
    "calories_kcal",
    "workout_minutes",
    "caffeine_mg",
    "alcohol_units",
    "screen_time_min",
    "mindfulness_minutes"
  ),
  "Behavioral Characteristics"
)
```

```{r}
# Sleep fields

create_descriptive_table(
  df,
  c(
    "sleep_duration_hours",
    "sleep_efficiency",
    "sleep_latency_min",
    "wake_after_sleep_onset_min",
    "sleep_stage_rem_pct",
    "sleep_stage_deep_pct",
    "sleep_stage_light_pct"
  ),
  "Sleep Characteristics"
)
```

```{r}
# Outcome variable

create_descriptive_table(
  df,
  c(
    "stress_score"
  ),
  "Stress Score Summary"
)
```

#### Missing Values

```{r}
# Investigating missing values

missing_values <- df %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Missing_Values"
  ) %>%
  mutate(
    Percent_Missing = round(100 * Missing_Values / nrow(df), 2)
  ) %>%
  arrange(desc(Missing_Values)) %>%
  mutate(
    Missing_Values = scales::comma(Missing_Values)
  )

missing_values %>%
  kable(caption = "Missing Values by Variable") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )

```

#### Categorical Fields

```{r}
# Function to graph frequencies of categorical data

plot_categorical <- function(data, column) {

  plot_data <- data %>%
    count({{ column }}) %>%
    mutate(
      percent = round(100 * n / sum(n), 1),
      label = paste0(
        scales::comma(n),
        "\n(",
        percent,
        "%)"
      )
    )

  ggplot(
    plot_data,
    aes(
      x = reorder(as.character({{ column }}), -n),
      y = n
    )
  ) +
    geom_col() +
    geom_text(
      aes(label = label),
      vjust = -0.3,
      size = 3.5
    ) +
    scale_y_continuous(
      expand = expansion(mult = c(0, 0.15))
    ) +
    labs(
      title = paste(
        "Frequency of",
        tools::toTitleCase(
          gsub("_", " ", deparse(substitute(column)))
        )
      ),
      x = NULL,
      y = "Count"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(
        face = "bold",
        hjust = 0.5
      ),
      axis.text.x = element_text(
        angle = 30,
        hjust = 1
      )
    )
}
```

```{r}
# Apply graphing function

plot_categorical(df, mood)

plot_categorical(df, gender)

plot_categorical(df, region)

plot_categorical(df, workout_type)

plot_categorical(df, device_model)
```

#### Stress Distribution

```{r}
# Investigating distribution of stress score within the dataset

ggplot(df, aes(x = stress_score)) +
  geom_histogram(bins = 20) +
  labs(
    title = "Distribution of Stress Scores",
    subtitle = "Stress scores appear approximately normally distributed",
    x = "Stress Score",
    y = "Count"
  ) +
  theme_minimal()
```


### Modeling

-   Explain your data preprocessing and cleaning steps.

**Add modeling here**

### Results

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

**Add results discussion here**


### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

**Add conclusion here**

## References
