---
title: "Modeling Physiological and Behavioral Predictors of Stress Using a Linear Mixed Model on Wearable Health Data"
subtitle: "Using Synthetic Longitudinal Wearable Device Data"
author: "Savannah Davis<br>Bryce Mattos<br>Jessica Jones<br>Jordan Gainey<br>(Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"}


## Introduction

  Stress is a state where the emotional, psychological, or physiological systems of an individual experience a demand that exceeds their knowledge and ability to resolve. Whether perceived consciously or unconsciously, the body then responds by prompting organs and glands to produce hormones that will assist the individual in adapting to the situation. Normal physiological responses to acute stress are short-lived and allow the body to return to its natural state without any long-term consequences. However, chronic stress, which is the result of unresolved long-standing demands and challenges can result in physical complications and disease in the body [@can2022survey]. In fact, the effects of chronic stress are so serious that research in Europe showed stress as the second most severe work-related health issue [@can2022survey]. Insights on stress can of course be self-reported by individuals, however, more efficient and accurate means of tracking stress can be accomplished via wearable devices such as Fitbit and Garmin.

	 
  Wearable devices can collect a wide range of data such as: heart rate, heart rate variability (HRV), sleep quality, physical activity, and movement patterns [@namvari2022ppg; @wang2024heart]. Out of those variables, researchers select the physiological ones that are directly influenced by the autonomic nervous system (ANS), mainly because stress is fundamentally an ANS response. The ANS regulates involuntary bodily functions and also consists of the sympathetic (SNS) and parasympathetic nervous systems (PNS) which maintain homeostasis within the body [@chalmers2022sleep]. During periods of stress, the SNS activates which results in elevated heart rates and changes in HRV, which makes those important variables to include in stress prediction research. Additionally, sleep, physical activity, circadian rhythms, and environmental conditions also contribute to stress and researchers often incorporate them when developing stress prediction models [@wang2024heart].
	
  
  While physiological variables such as heart rate and HRV provide valuable information about how the body responds to stress, researchers have found that behavioral and lifestyle factors also contribute to stress levels. Sleep duration, sleep quality, sleep efficiency, screen time, caffeine consumption, alcohol use, and mindfulness practices all have been shown to affect overall well-being [@vidalbustamante2024sleep; @booth2022stress]. Out of those listed factors, sleep has been identified as one of the strongest predictors of perceived stress, as disruptions influence an individual's ability to adapt to daily demands [@bloomfield2024sleep]. So, to increase the accuracy of stress prediction, researchers have concluded that combining physiological signals with lifestyle factors improves the ability to predict individuals experiencing higher levels of stress [@sano2018markers].


  Based on previous research, both physiological and behavioral measures appear to contribute to stress. This study builds on that work by examining these relationships using a synthetic longitudinal wearable dataset consisting of 300 users observed over a six-month period, resulting in more than 55,000 observations. The dataset includes physiological and behavioral measures such as heart rate variability (HRV), sleep duration, sleep efficiency, caffeine consumption, alcohol use, screen time, mindfulness, and physical activity, along with daily stress scores.


  The purpose of this study is to identify physiological and behavioral predictors of stress using longitudinal wearable health data and evaluate the ability of a linear mixed model to predict stress scores over time. Because the dataset contains repeated measurements from the same participants, a linear mixed model was selected to account for the correlation among observations collected from the same individual. Linear mixed models are well suited for longitudinal data because they account for repeated observations while capturing both population-level trends and variation between individuals [@silveira2023mixedeffects; @harrison2018mixed].


## Methods

A basic linear mixed model can be written as:

$$
y_{ij}
=
\beta_0
+
\beta_1X_{1ij}
+
\beta_2X_{2ij}
+
\cdots
+
\beta_pX_{pij}
+
b_{0i}
+
\varepsilon_{ij}
$$
where:

- $y_{ij}$ = observed stress score for participant $i$ on day $j$
- $\beta_0$ = fixed intercept
- $\beta_k$ = fixed-effect coefficient for predictor $k$
- $X_{kij}$ = observed value of predictor $k$ for participant $i$ on day $j$
- $b_{0i}$ = participant-specific random intercept
- $\varepsilon_{ij}$ = residual error


## Analysis and Results

### Data Exploration and Visualization

This report utilizes the *Health + Wellness + Stress/Sleep Tracking* dataset obtained from Kaggle. The dataset contains wearable health measurements and behavioral measurements for 300 simulated individuals over a 6 month period. The dataset contains demographic, physiological, behavioral, and sleep-related variables collected daily from wearable devices. A linear mixed model is appropriate for this dataset because observations from the same participant are not independent, violating a key assumption of ordinary linear regression. [@kaggle_dataset]

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(corrplot)
library(lme4)
library(lmerTest)
library(performance)
library(broom.mixed)
library(car)
```

```{r}
# Load Data

df = read_csv("wearables_health_6mo_daily.csv")
```

```{r}
# Quick overview of dataset

dataset_overview <- tibble(
  Metric = c("Total Observations", "Total Variables", "Unique Users"),
  Value = c(
    scales::comma(nrow(df)),
    scales::comma(ncol(df)),
    scales::comma(n_distinct(df$user_id))
  )
)

dataset_overview %>%
  kable(caption = "Table 1. Dataset Overview") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

The raw dataset contains over 55k records for 300 distinct users, with over 30 variables to choose to build the model from.

#### Descriptive Statistics

```{r}
# Rename variables to readable labels
variable_labels <- c(
  age = "Age",
  height_cm = "Height (cm)",
  weight_kg = "Weight (kg)",
  gender = "Gender",
  bmi = "BMI",
  resting_hr_bpm = "Resting Heart Rate",
  avg_hr_day_bpm = "Avg Heart Rate",
  hrv_rmssd_ms = "HRV (RMSSD)",
  spo2_avg_pct = "Average SpO₂ (%)",
  sbp_mmHg = "Systolic Blood Pressure",
  dbp_mmHg = "Diastolic Blood Pressure",
  steps = "Steps",
  distance_km = "Distance (km)",
  calories_kcal = "Calories",
  workout_minutes = "Workout Duration (min)",
  caffeine_mg = "Caffeine Intake (mg)",
  alcohol_units = "Alcohol Units",
  screen_time_min = "Screen Time (min)",
  mindfulness_minutes = "Mindfulness (min)",
  sleep_duration_hours = "Sleep Duration (hours)",
  sleep_efficiency = "Sleep Efficiency",
  sleep_latency_min = "Sleep Latency (min)",
  wake_after_sleep_onset_min = "Wake After Sleep Onset (min)",
  sleep_stage_rem_pct = "REM Sleep (%)",
  sleep_stage_deep_pct = "Deep Sleep (%)",
  sleep_stage_light_pct = "Light Sleep (%)",
  stress_score = "Stress Score"
)

# Function to create descriptive statistics tables

create_descriptive_table <- function(data, variables, caption) {

  data %>%
    select(all_of(variables)) %>%
    pivot_longer(
      cols = everything(),
      names_to = "Variable",
      values_to = "Value"
    ) %>%
    group_by(Variable) %>%
    summarise(
      Mean = round(mean(Value, na.rm = TRUE), 2),
      SD = round(sd(Value, na.rm = TRUE), 2),
      Min = round(min(Value, na.rm = TRUE), 2),
      Median = round(median(Value, na.rm = TRUE), 2),
      Max = round(max(Value, na.rm = TRUE), 2),
      .groups = "drop"
    ) %>%
        mutate(
      Variable = ifelse(
        Variable %in% names(variable_labels),
        variable_labels[Variable],
        Variable
      )
    ) %>%
    kable(caption = caption) %>%
    kable_styling(
      bootstrap_options = c(
        "striped",
        "hover",
        "condensed"
      ),
      full_width = FALSE
    )
}
```

The variables in the dataset are split into several different categories to assist in choosing variables of interest. Descriptive statistics tables will be created for each category including the mean, standard deviation, 
minimum, median, and maximum values.

```{r}
# Demographic fields

create_descriptive_table(
  df,
  c(
    "age",
    "height_cm",
    "weight_kg",
    "bmi"
  ),
  "Table 2. Demographic Characteristics"
)
```


```{r}
# Physiological fields

create_descriptive_table(
  df,
  c(
    "resting_hr_bpm",
    "avg_hr_day_bpm",
    "hrv_rmssd_ms",
    "spo2_avg_pct",
    "sbp_mmHg",
    "dbp_mmHg"
  ),
  "Table 3. Physiological Characteristics"
)
```

```{r}
# Behavioral fields

create_descriptive_table(
  df,
  c(
    "steps",
    "distance_km",
    "calories_kcal",
    "workout_minutes",
    "caffeine_mg",
    "alcohol_units",
    "screen_time_min",
    "mindfulness_minutes"
  ),
  "Table 4. Behavioral Characteristics"
)
```

```{r}
# Sleep fields

create_descriptive_table(
  df,
  c(
    "sleep_duration_hours",
    "sleep_efficiency",
    "sleep_latency_min",
    "wake_after_sleep_onset_min",
    "sleep_stage_rem_pct",
    "sleep_stage_deep_pct",
    "sleep_stage_light_pct"
  ),
  "Table 5. Sleep Characteristics"
)
```

```{r}
# Outcome variable

create_descriptive_table(
  df,
  c(
    "stress_score"
  ),
  "Table 6. Stress Score Summary"
)
```

#### Missing Values

The dataset is investigated for missing values to assess next steps for preprocessing.

```{r}
# Investigating missing values

missing_values <- df %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Missing_Values"
  ) %>%
  mutate(
    Percent_Missing = round(100 * Missing_Values / nrow(df), 2)
  ) %>%
  filter(
    Percent_Missing > 0
  ) %>%
  arrange(desc(Missing_Values)) %>%
  mutate(
    Variable = ifelse(
      Variable %in% names(variable_labels),
      variable_labels[Variable],
      Variable
    ),
    Missing_Values = scales::comma(Missing_Values)
  )

missing_values %>%
  kable(caption = "Table 7. Missing Values by Variable") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

```{r}
# Count rows containing at least one missing value

rows_with_missing <- sum(!complete.cases(df))

# Calculate percentage of dataset removed

pct_rows_removed <- round(
  100 * rows_with_missing / nrow(df),
  2
)

missing_summary <- tibble(
  Total_Observations = scales::comma(nrow(df)),
  Rows_With_Missing = scales::comma(rows_with_missing),
  Percent_Removed = paste0(pct_rows_removed, "%")
)

missing_summary %>%
  kable(
    caption = "Table 8. Impact of Missing Data Removal"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

Table 8 reveals that less than 2% of the observations contained at least one missing value, with missing values occurring across several physiological, behavioral, and sleep-related variables. Because only 1.91% of the observations contained missing data, a complete-case analysis was performed by removing incomplete observations prior to model fitting. This resulted in a relatively small loss of data while simplifying the subsequent analyses.

#### Stress Distribution

Before building the model, the distribution of the outcome variable is examined. Looking at the distribution of stress scores provides a better understanding of the data and helps identify any obvious issues such as extreme skewness or outliers.

```{r}
# Investigating distribution of stress score within the dataset

ggplot(df, aes(x = stress_score)) +
  geom_histogram(bins = 20) +
  labs(
    title = "Figure 1. Distribution of Stress Scores",
    subtitle = "Stress scores appear approximately normally distributed",
    x = "Stress Score",
    y = "Count"
  ) +
  theme_minimal()
```

Figure 1 shows that the distribution appears to be approximately normally distributed with no extreme outliers.

### Modeling and Results


```{r}
# Remove null values since there aren't many
df_clean <- na.omit(df)

# Convert categorical data to factor
df_clean <- df_clean %>%
  mutate(
    user_id = as.factor(user_id),
    gender = as.factor(gender),
    region = as.factor(region),
    device_model = as.factor(device_model),
    workout_type = as.factor(workout_type),
    mood = as.factor(mood)
  )

# Select all categorical variables
categorical_vars <- df_clean %>%
  select(
    user_id,
    gender,
    region,
    device_model,
    workout_type,
    mood
  )

# Select all continuous variables
continuous_vars <- df_clean %>%
  select(
    age, height_cm, weight_kg, bmi, resting_hr_bpm,
    avg_hr_day_bpm, hrv_rmssd_ms, spo2_avg_pct, sbp_mmHg,
    dbp_mmHg, steps, distance_km, calories_kcal,
    workout_minutes, caffeine_mg, alcohol_units, screen_time_min, 
    sleep_duration_hours, sleep_efficiency, sleep_latency_min,
    wake_after_sleep_onset_min, sleep_stage_rem_pct, sleep_stage_deep_pct, 
    sleep_stage_light_pct, mindfulness_minutes
  )
```

Before building the model, the continuous variables are checked for correlation. Highly correlated predictors can provide overlapping information, making it difficult to determine the individual effect of each variable. The heatmap provides a quick way to identify variables that may need to be removed before fitting the model.

```{r}
# Calculate correlation matrix for continuous variables
# Helps identify correlated variables before building model
continuous_vars_plot <- continuous_vars

colnames(continuous_vars_plot) <- ifelse(
  colnames(continuous_vars_plot) %in% names(variable_labels),
  variable_labels[colnames(continuous_vars_plot)],
  colnames(continuous_vars_plot)
)

# Correlation matrix
corr_matrix <- cor(
  continuous_vars_plot,
  use = "complete.obs"
)

# Correlation heatmap
corrplot(
  corr_matrix,
  method = "color",
  type = "upper",
  tl.cex = 0.7,
  title = "Figure 2. Correlation Heatmap of Continuous Predictor Variables",
  mar = c(0, 0, 2, 0)
)
```


```{r}
# Create table from the corr matrix
corr_table <- as.data.frame(as.table(corr_matrix)) %>%
  rename(
    Variable1 = Var1,
    Variable2 = Var2,
    Correlation = Freq
  ) %>%
  filter(
    as.character(Variable1) < as.character(Variable2)
  ) %>%
  mutate(
    abs_corr = abs(Correlation)
  ) %>%
  arrange(desc(abs_corr))

# Filter to only the highly correlated pairs and remove duplicate pairs
corr_table %>%
  filter(abs_corr >= 0.70) %>%
  mutate(
    Correlation = round(Correlation, 2)
  ) %>%
  select(
    Variable1,
    Variable2,
    Correlation
  ) %>%
  kable(
    caption = "Table 9. Strong correlations among predictor variables"
  ) %>%
  kable_styling(full_width = FALSE)
```

As expected, steps and distance were highly correlated because both measure daily physical activity. Similarly, BMI and weight were strongly correlated because BMI is calculated directly from weight and height. To reduce redundancy and minimize multicollinearity, only one variable from each highly correlated pair was retained for model development.


The full linear mixed model was fit using all remaining variables after removing the highly correlated predictors identified in the correlation analysis. The initial model gives a starting point for determining which variables contribute the most to predicting stress and which variables can be removed to simplify the model.

```{r}
# Build initial model with all variables (except ones with strong correlations)
# Put user_id as the random effect

lmm_full <- lmer(
  stress_score ~
    age +
    gender +
    region +
    device_model +
    workout_type +
    bmi +
    resting_hr_bpm +
    hrv_rmssd_ms +
    spo2_avg_pct +
    sbp_mmHg +
    dbp_mmHg +
    distance_km +
    workout_minutes +
    caffeine_mg +
    alcohol_units +
    screen_time_min +
    mindfulness_minutes +
    sleep_duration_hours +
    sleep_efficiency +
    sleep_latency_min +
    wake_after_sleep_onset_min +
    (1 | user_id),
  data = df_clean
)

#summary(lmm_full)
```

The initial model identified several physiological, behavioral, and sleep-related variables that demonstrated statistically significant associations with stress score. In contrast, demographic characteristics, device-related variables, resting heart rate, blood pressure, oxygen saturation, workout duration, and sleep latency did not demonstrate statistically significant relationships after accounting for the remaining predictors.


```{r}
# Reduced model after getting rid of irrelevant variables
# Variables with weak effects and little explanatory value were removed

lmm_final <- lmer(
  stress_score ~
    workout_type +
    hrv_rmssd_ms +
    distance_km +
    caffeine_mg +
    alcohol_units +
    screen_time_min +
    mindfulness_minutes +
    sleep_duration_hours +
    sleep_efficiency +
    wake_after_sleep_onset_min +
    (1 | user_id),
  data = df_clean
)

#summary(lmm_final)
```

Variables that demonstrated limited contribution to the interpretation of the initial model were removed to produce a more parsimonious model while retaining the primary wearable health measures of interest. The resulting model retained workout type, HRV, distance traveled, caffeine intake, alcohol consumption, screen time, mindfulness minutes, sleep duration, sleep efficiency, and wake after sleep onset.

After removing the additional predictors, the remaining coefficient estimates and significance levels were similar to those observed in the initial model. This suggests that the primary relationships identified in the larger model were preserved while reducing the number of predictors included in the analysis.

The reduced model was selected as the final model because it retained the principal wearable health variables while providing a simpler model for interpretation.

```{r}
# This table summarizes the statistical significance of each predictor retained in the final model
coef_table <- broom.mixed::tidy(lmm_final)

coef_table %>%
  filter(effect == "fixed") %>%
  select(
    term,
    estimate,
    std.error,
    statistic,
    p.value
  ) %>%
  mutate(
    term = case_when(
      term == "(Intercept)" ~ "Intercept",
      grepl("^workout_type", term) ~
        paste0("Workout Type: ", sub("^workout_type", "", term)),
      TRUE ~ ifelse(
        term %in% names(variable_labels),
        variable_labels[term],
        term
      )
    ),
    across(
      c(estimate, std.error, statistic),
      ~ round(.x, 3)
    ),
    p.value = ifelse(
      p.value < .001,
      "<0.001",
      round(p.value, 3)
    )
  ) %>%
  kable(
    col.names = c(
      "Variable",
      "Estimate",
      "Std. Error",
      "t Statistic",
      "p-value"
    ),
    caption = "Table 10. Fixed Effects from Final Linear Mixed Model"
  ) %>%
  kable_styling(
    bootstrap_options = c(
      "striped",
      "hover",
      "condensed"
    ),
    full_width = FALSE
  )
```

Table 10 summarizes the fixed effects from the final linear mixed model. The estimates show the average relationship between each predictor and stress score after accounting for the repeated measurements collected from each participant.

Several physiological and behavioral variables were significant predictors of stress. Higher heart rate variability (HRV), greater distance traveled, more mindfulness minutes, longer sleep duration, and higher sleep efficiency were all associated with lower stress scores. In contrast, greater caffeine intake, alcohol consumption, and screen time were associated with higher stress scores. These findings suggest that healthier sleep patterns, higher HRV, regular physical activity, and mindfulness are associated with lower levels of stress, while greater caffeine consumption, alcohol use, and screen time are associated with higher stress levels.

None of the workout type indicator variables were statistically significant after accounting for the remaining predictors. Although participants who reported running tended to have slightly lower stress scores than the reference group, the estimated effect was small and only marginally significant. Overall, the physiological, behavioral, and sleep-related variables included in the model appeared to have a stronger relationship with stress score than the reported workout type.


The final fitted model is:

$$
\begin{aligned}
\widehat{\text{Stress Score}}_{ij}
&= 116.1
-0.135(\text{Workout}_{\text{Mixed}})
+0.084(\text{Workout}_{\text{None}})
-0.195(\text{Workout}_{\text{Run}})\\
&\quad
+0.093(\text{Workout}_{\text{Strength}})
-0.169(\text{Workout}_{\text{Walk}})
+0.153(\text{Workout}_{\text{Yoga}})\\
&\quad
-0.258(\text{HRV})
-0.442(\text{Distance})
+0.0396(\text{Caffeine})
+4.577(\text{Alcohol})\\
&\quad
+0.0324(\text{Screen Time})
-0.0417(\text{Mindfulness})
-5.974(\text{Sleep Duration})\\
&\quad
-9.564(\text{Sleep Efficiency})
+0.00520(\text{Wake After Sleep Onset})
+b_{0i}.
\end{aligned}
$$
Here, $i$ indexes participants, $j$ indexes daily observations, and $b_{0i}$ represents the participant-specific random intercept.

##### Figure 3. Normality of Residuals

```{r}
# Check model diagnostics
check_model(lmm_final, check = "qq")
```

##### Figure 4. Homogeneity of Variance

```{r}
# Check model diagnostics
check_model(lmm_final, check = "homogeneity")
```

##### Figure 5. Linearity

```{r}
# Check model diagnostics
check_model(lmm_final, check = "linearity")
```

The assumptions of the final linear mixed model were checked. Figure 3 above shows that the residuals are approximately normally distributed, with only small departures from normality at the extreme ends of the distribution. In Figure 4, the homogeneity of variance plot shows a fairly consistent spread of the residuals across the fitted values, with no strong evidence of increasing or decreasing variance. In Figure 5, the residuals versus fitted values plot also does not show any obvious patterns, indicating that the relationship between the predictors and stress score is approximately linear. Overall, these diagnostic plots suggest that the assumptions of the linear mixed model were reasonably satisfied.

```{r}
# Calculate model performance metrics

r2_vals <- performance::r2(lmm_final)

icc_val <- performance::icc(lmm_final)

# Residuals from final model
model_residuals <- resid(lmm_final)

# Error metrics
rmse <- sqrt(mean(model_residuals^2))

mae <- mean(abs(model_residuals))

performance_table <- tibble(
  Metric = c(
    "Marginal R²",
    "Conditional R²",
    "ICC",
    "RMSE",
    "MAE",
    "Users",
    "Observations"
  ),
  Value = c(
    round(r2_vals$R2_marginal, 3),
    round(r2_vals$R2_conditional, 3),
    round(icc_val$ICC_adjusted, 3),
    round(rmse, 3),
    round(mae, 3),
    scales::comma(n_distinct(df_clean$user_id)),
    scales::comma(nrow(df_clean))
  )
)

performance_table %>%
  kable(
    caption = "Table 11. Final Model Performance Metrics"
  ) %>%
  kable_styling(
    bootstrap_options = c(
      "striped",
      "hover",
      "condensed"
    ),
    full_width = FALSE
  )
```

Table 11 above summarizes the overall performance of the final linear mixed model. The marginal $R^2$ of 0.726 indicates that the fixed effects explained approximately 72.6% of the variation in stress scores. The conditional $R^2$ was nearly identical (0.729), indicating that the participant-specific random intercept contributed relatively little additional explained variation beyond the fixed effects. Similarly, the ICC of 0.013 suggests that only a small proportion of the total variability in stress scores was attributable to stable differences between participants.

The RMSE of 6.072 indicates a typical residual error of approximately six stress-score points, with larger prediction errors receiving greater weight. The MAE of 4.884 indicates that the fitted stress scores differed from the observed values by an average of approximately five points. 

Collectively, these performance metrics indicate that the final model explained a substantial proportion of the variation in stress scores while maintaining relatively small residual errors across the observed data.

```{r}
# Observed versus fitted values

ggplot(
  data.frame(
    Observed = df_clean$stress_score,
    Fitted = fitted(lmm_final)
  ),
  aes(
    x = Observed,
    y = Fitted
  )
) +
  geom_point(
    alpha = 0.05,
    size = 1
  ) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed"
  ) +
  labs(
    title = "Figure 6. Observed versus Fitted Stress Scores",
    x = "Observed Stress Score",
    y = "Fitted Stress Score"
  ) +
  theme_minimal()
```

Figure 6 above compares the observed stress scores with the fitted values produced by the final linear mixed model. Most observations fall relatively close to the reference line, indicating that the fitted values generally correspond well with the observed stress scores. Although some variability remains, the overall pattern suggests that the model provides a good representation of the observed data.


### Conclusion

Overall, the final linear mixed model identified several physiological, behavioral, and sleep-related variables that were associated with stress within this synthetic wearable health dataset. In particular, measures related to heart rate variability, physical activity, sleep quality, mindfulness, caffeine intake, alcohol consumption, and screen time were significantly associated with stress after accounting for repeated measurements within participants. The final model satisfied the evaluated assumptions, explained a substantial proportion of the variation in stress scores, and provided a good overall fit to the observed data.

These findings demonstrate the value of linear mixed models for analyzing longitudinal wearable health data, where repeated observations from the same individuals violate the assumption of independent observations required by traditional linear regression. By incorporating participant-specific random effects, the model was able to account for within-participant correlation while estimating the relationships between wearable health measures and stress.

Because this study used a synthetic dataset, the findings should not be interpreted as evidence of real-world clinical relationships. However, the analysis illustrates how wearable health data and linear mixed modeling can be combined to investigate factors associated with stress and provides a framework that could be applied to real-world longitudinal health data in future research.

## References
