---
title: "Modeling Physiological and Behavioral Predictors of Stress Using a Linear Mixed Model on Wearable Health Data"
subtitle: "Using Synthetic Longitudinal Wearable Device Data"
author: "Savannah Davis<br>Bryce Mattos<br>Jessica Jones<br>Jordan Gainey<br>(Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  revealjs:
    theme: simple
    width: 1600
    height: 900
    margin: 0.06
    slide-number: true
    transition: fade
    incremental: false
    code-fold: true
    code-summary: "Show code"
    footer: "Modeling Stress from Wearable Health Data - LMM"
execute:
  echo: false
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
---


```{r}
#| label: setup
#| include: false

library(tidyverse)
library(knitr)
library(kableExtra)
library(corrplot)
library(lme4)
library(lmerTest)
library(performance)
library(broom.mixed)

df <- read_csv("wearables_health_6mo_daily.csv", show_col_types = FALSE)

# Summarize rows containing missing values before cleaning

rows_with_missing <- sum(!complete.cases(df))

pct_rows_removed <- round(
  100 * rows_with_missing / nrow(df),
  2
)

complete_observations <- sum(complete.cases(df))

# Remove incomplete observations and format variables

df_clean <- df %>%
  drop_na() %>%
  mutate(
    across(
      c(gender, region, device_model, workout_type),
      as.factor
    ),
    user_id = as.factor(user_id)
  )

variable_labels <- c(
  age = "Age",
  height_cm = "Height (cm)",
  weight_kg = "Weight (kg)",
  bmi = "BMI",
  resting_hr_bpm = "Resting Heart Rate",
  avg_hr_day_bpm = "Avg Heart Rate",
  hrv_rmssd_ms = "HRV (RMSSD)",
  spo2_avg_pct = "Average SpO₂ (%)",
  sbp_mmHg = "Systolic Blood Pressure",
  dbp_mmHg = "Diastolic Blood Pressure",
  steps = "Steps",
  distance_km = "Distance (km)",
  calories_kcal = "Calories",
  workout_minutes = "Workout Duration (min)",
  caffeine_mg = "Caffeine (mg)",
  alcohol_units = "Alcohol (units)",
  screen_time_min = "Screen time (min)",
  mindfulness_minutes = "Mindfulness (min)",
  sleep_duration_hours = "Sleep duration (hours)",
  sleep_efficiency = "Sleep efficiency",
  sleep_latency_min = "Sleep Latency (min)",
  wake_after_sleep_onset_min = "Wake after sleep onset (min)",
  sleep_stage_rem_pct = "REM Sleep (%)",
  sleep_stage_deep_pct = "Deep Sleep (%)",
  sleep_stage_light_pct = "Light Sleep (%)"
)

continuous_vars <- df_clean %>%
  select(
    age, height_cm, weight_kg, bmi, resting_hr_bpm,
    avg_hr_day_bpm, hrv_rmssd_ms, spo2_avg_pct, sbp_mmHg,
    dbp_mmHg, steps, distance_km, calories_kcal,
    workout_minutes, caffeine_mg, alcohol_units, screen_time_min,
    sleep_duration_hours, sleep_efficiency, sleep_latency_min,
    wake_after_sleep_onset_min, sleep_stage_rem_pct, sleep_stage_deep_pct,
    sleep_stage_light_pct, mindfulness_minutes
  )

lmm_final <- lmer(
  stress_score ~ 
    workout_type + 
    hrv_rmssd_ms + 
    distance_km +
    caffeine_mg + 
    alcohol_units + 
    screen_time_min +
    mindfulness_minutes + 
    sleep_duration_hours + 
    sleep_efficiency +
    wake_after_sleep_onset_min + 
    (1 | user_id),
  data = df_clean
)
```

# Introduction {background-color="#176B87" background-transition="fade" .section-slide}

## Why Study Stress?

- Chronic stress can contribute to serious physical and psychological health
  consequences ^[@can2022survey]
- Self-reported stress is useful, but it is intermittent and subjective
- Wearable devices continuously capture physiological and behavioral signals
  that may help identify changes associated with stress

Wearable stress research commonly uses physiological signals influenced by the
autonomic nervous system, including heart rate and HRV
^[@namvari2022ppg; @wang2024heart; @chalmers2022sleep]


## Why Wearable Data? {.smaller}

Wearables can capture multiple dimensions of daily life:

| Physiological | Behavioral | Sleep |
|---|---|---|
| Heart rate variability | Physical activity | Duration |
| Heart rate | Caffeine and alcohol | Efficiency |
| Oxygen saturation | Screen time | Wake after sleep onset |
| Blood pressure | Mindfulness | Sleep latency |

<br>

Stress is multifaceted, so combining these measures may be more informative
than using a single signal ^[@sano2018markers; @neigel2024wearables]

Sleep, activity, environmental conditions, caffeine, alcohol, screen time, and
mindfulness may also contribute to stress and well-being
^[@vankraaij2020chronic; @vidalbustamante2024sleep; @booth2022stress]

Sleep has been identified as an especially important predictor of perceived
stress in longitudinal wearable research ^[@bloomfield2024sleep]

## Research Objective


Identify physiological and behavioral predictors of stress and evaluate how
well a linear mixed model explains daily stress scores.

We focused on two questions:

1. Which wearable, behavioral, and sleep measures are associated with stress?
2. How well does the final model fit the observed stress scores?

# Methods {background-color="#176B87" background-transition="fade" .section-slide}

## Why a Linear Mixed Model (LMM)? {.smaller}

Each participant contributed many daily observations.

::: {.columns}
::: {.column width="48%"}
### Standard regression

- Assumes observations are independent
- Would treat every daily record as unrelated
:::

::: {.column width="4%"}
:::

::: {.column width="48%"}
### Linear mixed model

- Recognizes repeated days within participants
- Estimates population effects
- Allows participant-specific baselines
:::
:::

A standard linear regression assumes independent observations; that assumption
is not appropriate for this repeated-measures structure
^[@detry2016repeated; @murphy2022accessible]

LMMs are designed to estimate population-level relationships while accounting for observations nested within
participants ^[@harrison2018mixed; @silveira2023mixedeffects]

## Model Structure

$$
y_{ij}
=
\beta_0
+
\beta_1X_{1ij}
+
\beta_2X_{2ij}
+
\cdots
+
\beta_pX_{pij}
+
b_{0i}
+
\varepsilon_{ij}
$$

where:

- $y_{ij}$ = observed stress score for participant $i$ on day $j$
- $\beta_0$ = fixed intercept
- $\beta_k$ = fixed-effect coefficient for predictor $k$
- $X_{kij}$ = observed value of predictor $k$ for participant $i$ on day $j$
- $b_{0i}$ = participant-specific random intercept
- $\varepsilon_{ij}$ = residual error

The random intercept allows each participant to have a different baseline
stress level.

# Data Exploration and Visualization {background-color="#176B87" background-transition="fade" .section-slide}

## Dataset

```{r}
#| label: dataset-overview

tibble(
  `Participants` = n_distinct(df$user_id),
  `Daily observations` = scales::comma(nrow(df)),
  `Study period` = "Approximately 6 months",
  `Data type` = "Synthetic longitudinal data"
) %>%
  kable(align = "c") %>%
  kable_styling(full_width = TRUE, font_size = 36)
```

<br>

The dataset contains demographic, physiological, activity, lifestyle, sleep,
and daily stress measures for simulated participants ^[@kaggle_dataset]

Public wearable datasets such as WESAD have supported the development and evaluation
of physiological stress-detection methods ^[@schmidt2018wesad]

## Data Preparation

- Reviewed the dataset structure, variable types, and missingness
- Removed the small number of rows containing missing values
- Converted categorical variables to factors
- Summarized demographic, physiological, behavioral, sleep, and outcome
  variables
- Examined the distribution of stress scores before modeling

## Missing Values

```{r}
#| echo: false
#| warning: false
#| message: false
#| tbl-cap: "Impact of Missing Data"

missing_summary <- tibble(
  `Total Observations` = scales::comma(nrow(df)),
  `Incomplete Observations` = scales::comma(rows_with_missing),
  `Complete Observations` = scales::comma(nrow(df_clean)),
  `Percent Removed` = paste0(pct_rows_removed, "%")
)

missing_summary %>%
  kable(
    align = rep("c", 4)
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "condensed"),
    full_width = FALSE,
    position = "center"
  )
```

- Missing values occurred across several physiological, activity, and sleep-related variables
- Each affected variable contained 1,057 missing values, representing less than 2% of the original observations
- A complete-case analysis was performed by removing incomplete observations before model fitting
- This resulted in a relatively small loss of data

## Distribution of Stress Scores

```{r}
#| label: stress-distribution
#| fig-height: 5.4
#| fig-width: 9

ggplot(df_clean, aes(stress_score)) +
  geom_histogram(bins = 20) +
  labs(
    title = "Distribution of Stress Scores",
    subtitle = "Stress scores appear approximately normally distributed",
    x = "Stress Score",
    y = "Count"
  ) +
  theme_minimal()
```

Stress scores were approximately normally distributed, with most observations concentrated near the center of the scale and no obvious extreme outliers.


# Modeling and Results {background-color="#176B87" background-transition="fade" .section-slide}

## Model Development

1. Reviewed continuous predictors for strong correlations
2. Retained one variable from redundant pairs, including distance rather than
   steps and BMI rather than weight
3. Fit a full random-intercept model
4. Removed variables with limited contribution to create a more interpretable
   final model

## Correlation Heatmap

```{r}
#| label: correlation-heatmap
#| fig-height: 6.3
#| fig-width: 11

continuous_vars_plot <- continuous_vars

colnames(continuous_vars_plot) <- ifelse(
  colnames(continuous_vars_plot) %in% names(variable_labels),
  variable_labels[colnames(continuous_vars_plot)],
  colnames(continuous_vars_plot)
)

corr_matrix <- cor(continuous_vars_plot, use = "complete.obs")

corrplot(
  corr_matrix,
  method = "color",
  type = "upper",
  tl.cex = 0.7,
  title = "Correlation Heatmap of Continuous Predictor Variables",
  mar = c(0, 0, 2, 0)
)
```

## Correlation Review

- Continuous predictors were examined for strong pairwise correlations
- Steps and distance were highly correlated because both measured daily
  physical activity
- BMI and weight were highly correlated because BMI is derived from weight and
  height
- Distance and BMI were carried forward to the full model because they were easier to interpret than their correlated alternatives

Removing redundant predictors reduced multicollinearity and made individual
coefficients easier to interpret.

## Model Selection

1. A full random-intercept model was fitted using the remaining candidate predictors

2. Predictor contribution was evaluated based on statistical evidence and practical interpretability

3. Eleven variables with limited contribution were removed:

   - **Demographic/device:** age, gender, region, and device model
   - **Physiological:** BMI, resting heart rate, oxygen saturation, and systolic and diastolic blood pressure
   - **Behavioral/sleep:** workout duration and sleep latency

4. Workout type was retained because of its conceptual relevance to wearable activity, despite not being statistically significant


## Final Model

```{r}
#| label: final-model-list

tibble(
  Domain = c("Activity", "Physiological", "Behavioral", "Sleep"),
  Predictors = c(
    "Workout type; distance traveled",
    "Heart rate variability (HRV)",
    "Caffeine; alcohol; screen time; mindfulness",
    "Duration; efficiency; wake after sleep onset"
  )
) %>%
  kable(align = c("l", "l")) %>%
  kable_styling(full_width = TRUE, font_size = 36)
```

<br>

All predictors were modeled as fixed effects, with a random intercept for
participant. 

The final model was estimated using restricted maximum likelihood.

## Final Fitted Model

The final fitted model can be written as:

$$
\begin{aligned}
\widehat{\text{Stress Score}}_{ij}
&= 116.118
-0.135(\text{Workout}_{\text{Mixed}})
+0.084(\text{Workout}_{\text{None}})
-0.195(\text{Workout}_{\text{Run}})\\
&\quad
+0.093(\text{Workout}_{\text{Strength}})
-0.169(\text{Workout}_{\text{Walk}})
+0.153(\text{Workout}_{\text{Yoga}})\\
&\quad
-0.258(\text{HRV})
-0.442(\text{Distance})
+0.040(\text{Caffeine})
+4.577(\text{Alcohol})\\
&\quad
+0.032(\text{Screen Time})
-0.042(\text{Mindfulness})
-5.974(\text{Sleep Duration})\\
&\quad
-9.564(\text{Sleep Efficiency})
+0.005(\text{Wake After Sleep Onset})
+b_{0i}.
\end{aligned}
$$

where:

- $i$ indexes participants
- $j$ indexes daily observations
- $b_{0i}$ represents the participant-specific random intercept

## Fixed Effects: Workout Type {.smaller}

```{r}
#| label: fixed-effects-workout

coef_table <- broom.mixed::tidy(
  lmm_final,
  effects = "fixed",
  conf.int = TRUE,
  conf.level = 0.95
) %>%
  mutate(
    Variable = case_when(
      term == "(Intercept)" ~ "Intercept",
      str_detect(term, "^workout_type") ~
        paste0("Workout Type: ", str_remove(term, "^workout_type")),
      term %in% names(variable_labels) ~ unname(variable_labels[term]),
      TRUE ~ term
    ),
    Coefficient = round(estimate, 3),
    `95% Confidence Interval` = paste0(
      "(", round(conf.low, 4), ", ", round(conf.high, 4), ")"
    ),
    `p-value` = if_else(
      p.value < .001,
      "<0.001",
      format(round(p.value, 4), nsmall = 4)
    )
  )

coef_table %>%
  filter(term == "(Intercept)" | str_detect(term, "^workout_type")) %>%
  select(Variable, Coefficient, `95% Confidence Interval`, `p-value`) %>%
  kable(
    caption = "Fixed Effects from Final LMM: Intercept and Workout Type",
    align = c("l", "c", "c", "c")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "condensed"),
    full_width = TRUE,
    font_size = 28
  )
```

<br>

- None of the workout-type indicators were statistically significant
- Running showed a slightly lower estimated stress score than the reference
  workout category, but its confidence interval included zero
- Workout type remained in the model because the type of physical activity is
  substantively relevant to wearable-health research

## Fixed Effects: Daily Predictors

```{r}
#| label: fixed-effects-predictors

coef_table %>%
  filter(term != "(Intercept)", !str_detect(term, "^workout_type")) %>%
  select(Variable, Coefficient, `95% Confidence Interval`, `p-value`) %>%
  kable(
    caption = "Fixed Effects from Final LMM: Daily Physiological, Behavioral, and Sleep Predictors",
    align = c("l", "c", "c", "c")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "condensed"),
    full_width = TRUE,
    font_size = 28
  )
```


## What Was Associated With Lower Stress?

Holding the other variables constant, lower stress scores were associated with:

- Higher heart rate variability
- Greater distance traveled
- More mindfulness minutes
- Longer sleep duration
- Higher sleep efficiency


## What Was Associated With Higher Stress?

Holding the other variables constant, higher stress scores were associated with:

- Greater caffeine intake
- Greater alcohol consumption
- More screen time
- More wake time after sleep onset


## Model Performance

```{r}
#| label: model-performance

r2_vals <- performance::r2(lmm_final)
icc_val <- performance::icc(lmm_final)
model_residuals <- resid(lmm_final)

tibble(
  Metric = c("Marginal R²", "Conditional R²", "ICC", "RMSE", "MAE"),
  Value = c(
    r2_vals$R2_marginal,
    r2_vals$R2_conditional,
    icc_val$ICC_adjusted,
    sqrt(mean(model_residuals^2)),
    mean(abs(model_residuals))
  )
) %>%
  mutate(Value = round(Value, 3)) %>%
  kable(align = "c") %>%
  kable_styling(full_width = TRUE, font_size = 36)
```

## Interpreting Performance

- Fixed effects explained approximately 72.6% of variation in stress scores
- Conditional $R^2$ was only slightly higher, so the random intercept added
  modest explanatory power
- The ICC of 0.013 indicates little stable between-participant variation
- Predictions differed from observed scores by about 6.1 points (RMSE) or
  4.8 points on average (MAE)

Although the ICC was low, the LMM remained appropriate for the repeated-measures study design.


## Observed vs. Fitted Stress Scores

```{r}
#| label: observed-fitted
#| fig-height: 5.8
#| fig-width: 9

ggplot(
  data.frame(
    Observed = df_clean$stress_score,
    Fitted = fitted(lmm_final)
  ),
  aes(
    x = Observed,
    y = Fitted
  )
) +
  geom_point(
    alpha = 0.05,
    size = 1
  ) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed"
  ) +
  labs(
    title = "Observed versus Fitted Stress Scores",
    x = "Observed Stress Score",
    y = "Fitted Stress Score"
  ) +
  theme_minimal()
```

Most observations fall near the reference line, although residual variation
remains.

## Assumption Checks

Model diagnostics indicated:

- Residuals were approximately normally distributed, with minor tail departures
- Residual variance was reasonably consistent across fitted values
- Residual plots showed no strong nonlinearity
- Participant-specific random effects were approximately normally distributed

Overall, the evaluated assumptions were reasonably satisfied, supporting
interpretation of the fitted model ^[@yang2014adv]

## Key Takeaways

1. Physiological, behavioral, and sleep measures jointly explained a substantial
   portion of variation in daily stress scores
2. HRV, distance, mindfulness, sleep duration, and sleep efficiency were
   associated with lower stress
3. Caffeine, alcohol, screen time, and wake after sleep onset were associated
   with higher stress
4. An LMM appropriately accounted for repeated daily measurements within
   participants

## Limitations and Future Work

- The data was synthetic
- Associations do not establish causation
- The model did not include time dependence
- Future work should validate the approach using real-world wearable data

# Conclusion {background-color="#176B87" background-transition="fade" .section-slide}

## Conclusion

Within this synthetic dataset, physiological, behavioral, and sleep measures explained a substantial proportion of variation in daily stress scores.

<br> 

The results demonstrate how a linear mixed model can combine repeated wearable measurements while accounting for multiple observations from the same participant.

<br>

The final model offered strong explanatory performance within this synthetic
dataset and identified interpretable relationships that could guide future
research using real wearable-health data.

## Questions?

**Thank you**

## References {.smaller}

::: {#refs}
:::
